---
title: "HW_4"
author: "Devon Park"
date: "2026-02-19"
output: github_document
---

```{r, include = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r}
library(survival)
library(survminer)
library(broom)
library(knitr)
library(tidyverse)
```

# Question 1A
We continue to use the hwdata1.csv as in Homework 1 to study risk factors for the breast cancer recurrence. Three Cox proportional hazards regression models were fit


### Prep Data:
```{r}
hwdata1 = read_csv("hwdata1-1.csv") |> 
  mutate(
    event = censrec, 
    #Event indicator: 1 = recurrence, 0 = censored (based on codebook)
    
    hormone_grp = factor(hormone, 
                         levels = c(1, 2), 
                         labels = c("Hormone: Yes", "Hormone: No")) |>
                          #Hormone therapy groups (codebook: 1=Yes, 2=No)
  
        relevel(ref = "Hormone: No"),
        #Making sure that "No Hormone" is the reference group
  
    meno_status = factor(menopause,
                      levels = c(1, 2),
                      labels = c("Menopausal: Yes", "Menopausal: No")) |> 
      relevel(ref = "Menopausal: No")
      #Women who are not yet experiencing menopause are the reference group
  )
    
```


### Model 1: Covariate: Hormone Therapy Group

General Model: $\lambda(t|hormone) = \lambda_0e^{\beta_1*hormone}$

The Cox proportional hazards with the covariate of horomone therapy group is as follows (beta rounded to three decimal places):
$\lambda(t|hormone) = \lambda_0e^{0.364*hormone}$
```{r}
model_1_results = hwdata1 |> coxph(Surv(rectime, event) ~ hormone_grp, 
                      data = _) |> 
          tidy(exponentiate = TRUE, conf.int = TRUE)

#Clean Variable Names in Output
model_1_results = model_1_results |>
  mutate(
    term = case_when(
      term == "hormone_grpHormone: Yes" ~ "Hormone: Yes",
      TRUE ~ term
    )
  )


#Cleaned Up Output
model_1_results |>
  knitr::kable(
    digits = 3,
    col.names = c("Variable", "Hazard Ratio", "Std Error", "Z statistic", "p-value", "Lower 95% CI", "Upper 95% CI"),
    caption = ""
  )
```

### Model 2: Covariates: Hormone therapy groups, Age, Menopause status, Tumor size, and Number of nodes

General Model: $\lambda(t|X) = \lambda_0e^{\beta_1*hormone + \beta_2*age + \beta_3*menopausal + \beta_4*size + \beta_5*nodes}$

The Cox proportional hazards with the above covariates is as follows (betas rounded to three decimal places):
$\lambda(t|X) = \lambda_0e^{0.381*hormone -0.014*age -0.353*menopausal + 0.008*size + 0.052*nodes}$
```{r}
model_2_results = hwdata1 |> coxph(Surv(rectime, event) 
                           ~ hormone_grp + age + meno_status + size + nodes,
                           data = _) |> 
          tidy(exponentiate = TRUE, conf.int = TRUE)


#Clean Variable Names in Output
model_2_results = model_2_results |>
  mutate(
    term = case_when(
      term == "hormone_grpHormone: Yes" ~ "Hormone: Yes",
      term == "meno_statusMenopausal: Yes" ~ "Menopausal: Yes",
      term == "nodes" ~ "Nodes involved",
      term == "age" ~ "Age (years)",
      term == "size" ~ "Tumor size (mm)",
      TRUE ~ term
    )
  )


#Cleaned Up Output
model_2_results |>
  knitr::kable(
    digits = 3,
    col.names = c("Variable", "Hazard Ratio", "Std Error", "Z statistic", "p-value", "Lower 95% CI", "Upper 95% CI"),
    caption = ""
  )
```


### Model 3: Covariates: Hormone therapy groups, Age, Menopause status, Tumor size, Number of nodes, and the Interaction between hormone therapy groups and number of nodes

General Model: $\lambda(t|X) = \lambda_0e^{\beta_1*hormone + \beta_2*age + \beta_3*menopausal + \beta_4*size + \beta_5*nodes + \beta_6*nodes*hormone}$

The Cox proportional hazards model with interaction between hormone therapy and nodes is as follows (betas rounded to three decimal places): $\lambda(t|X) = \lambda_0e^{0.630*hormone  -0.015*age  -0.393*menopausal + 0.007*size + 0.083*nodes -0.038*nodes*hormone}$



```{r}
model_3_results = hwdata1 |> coxph(Surv(rectime, event) 
                           ~ hormone_grp*nodes + age + meno_status + size,
                           data = _)|> 
          tidy(exponentiate = TRUE, conf.int = TRUE)

model_3_results = model_3_results |>
  mutate(
    term = case_when(
      term == "hormone_grpHormone: Yes" ~ "Hormone: Yes",
      term == "hormone_grpHormone: Yes:nodes" ~ "Hormone: Yes × nodes",
      term == "meno_statusMenopausal: Yes" ~ "Menopausal: Yes",
      term == "nodes" ~ "Nodes involved",
      term == "age" ~ "Age (years)",
      term == "size" ~ "Tumor size (mm)",
      TRUE ~ term
    )
  )

#Cleaned Up Output
model_3_results |>
  knitr::kable(
    digits = 3,
    col.names = c("Variable", "Hazard Ratio", "Std Error", "Z statistic", "p-value", "Lower 95% CI", "Upper 95% CI"),
    caption = ""
  )



#hormone_grp*nodes --> includes the individual terms as well as the interaction term
```

Betas of all the models 
```{r}
#Beta for Model 1
model_1_betas = hwdata1 |> coxph(Surv(rectime, event) ~ hormone_grp, 
                      data = _) 
model_1_betas

#Betas for Model 2
model_2_betas = hwdata1 |> coxph(Surv(rectime, event) 
                           ~ hormone_grp + age + meno_status + size + nodes,
                           data = _)
model_2_betas

#Betas for Model 3

model_3_beta = hwdata1 |> coxph(Surv(rectime, event) 
                           ~ hormone_grp*nodes + age + meno_status + size,
                           data = _)

model_3_betas = model_3_beta|> 
  tidy(exponentiate = FALSE, conf.int = FALSE)


model_3_betas = model_3_betas |>
  mutate(
    term = case_when(
      term == "hormone_grpHormone: Yes" ~ "Hormone: Yes",
      term == "hormone_grpHormone: Yes:nodes" ~ "Hormone: Yes × nodes",
      term == "meno_statusMenopausal: Yes" ~ "Menopausal: Yes",
      term == "nodes" ~ "Nodes involved",
      term == "age" ~ "Age (years)",
      term == "size" ~ "Tumor size (mm)",
      TRUE ~ term
    )
  ) |> 
  knitr::kable(
    digits = 3,
    col.names = c("Variable", "Betas", "Std Error", "Z statistic", "p-value"),
    caption = ""
  )
model_3_betas
```


Get covariance matrix
```{r}
vcov(model_3_beta)
```


