---
title: "HW_3"
author: "Devon Park"
date: "2026-02-15"
output: github_document
---

```{r, include = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r}
library(survival)
library(survminer)
```

## Question 1 

The Framingham Heart Study has collected data on cardiovascular risk factors and long-term follow-up for nearly 5,000 residents of Framingham, Massachusetts. These data were used as an example in Section 7.5 of Dupont (2008).

An investigator wishes to study the association between baseline age and the risk of coronary heart disease (CHD) among 4,699 participants using a Cox proportional hazards regression model.


#### Question 1A 

(a) Specify the Cox proportional hazards model with baseline age included as a continuous covariate. Using SAS/R, estimate the hazards ratio and its 95% confidence interval.


```{r}
framingham = read.csv("HW3\framingham.csv")
```





